Divergent elliptical biface morphology: Distinct local retouch practice or serpent dance proxy for Elapids and Vipers?

### 2022-12-20 supp/index.qmd
<edit index>

Elliptical bifaces are prevalent at Mississippian and Caddo sites throughout the American Southeast, and are depicted in motifs that adorn shell gorgets and bowls, representing two distinct dances. Morphological attributes of elliptical bifaces used in Mississippian and Caddo imagery differs in noticeable ways that are manifest in associated material culture elements, and bifaces from Caddo sites more readily articulate with the serpent dance, where the motif portrays the emergence of the Hero Twins from the Beneath World; each grasping an elliptical biface.

@@

Among the shell cup illustrations assembled by Phillips and Brown [-@RN2447] that characterize the material culture at Spiro Mound is a depiction of elliptical bifaces conveyed as elements of the serpent/snake dance rather than elements of the eagle/bird dance. It is plausible that in this case biface **shape** may convey a relevant---and ceremonially important---representation related to extant morphological differences expressed in local elapid and viper species. Differences in biface **size** may articulate with the size of a specific species, as some are smaller than others (i.e. _Sistrurus miliarius streckeri_ or western pygmy rattlesnake). Alternatively, **size** differences might also depict allometric trends associated with the development and maturation of distinct species, as well as sexual size dimorphism.

@@

1.   Elliptical bifaces found at Millsap Cache were manufactured using Kay County Chert, while those from Jowell Farm were manufactured using Edwards Chert. Differences in stone tool shape as a function of raw material have been reported for Perdiz arrow points in the region [@RN9364]; however, others have demonstrated that the attribution of morphological variability to raw material differences is an assumption that warrants regular testing, suggesting that primacy should not be assumed [@RN8442]. Thus, a comparison of between-site morphology was included to assess elliptical biface morphology as a function of raw material.
  a)    The biogeography of elapid and viper species differs within the ancestral Caddo area, where elapid populations occur south of the Red River, excluding them from the geography surrounding Millsap Cache. If elliptical biface morphology was intended to represent local elapid and viper populations, then elliptical bifaces from these sites will differ due to differences in the constituent communities of elapids and vipers.

@@

In the case of ceremonial bifaces, the intentional use of parallel oblique flake scars may have been meant to convey motion, where flake scars on one face represent muscle constriction along one edge and muscle dilation on the other, where the pattern is inverted on the opposite face. Thus, in intentionally flipping between the two faces---a potential component of the serpent/snake dance---dancers could convey motion using a static implement. The parallel oblique flake scars closely mimic the parallel scales that occur beneath both elapids and vipers, which shift in similar ways to generate motion.

>>>>

Elliptical bifaces are prevalent at Mississippian and Caddo sites throughout the American Southeast, and are depicted in motifs that adorn shell gorgets and bowls, representing two distinct dances. Morphological attributes of elliptical bifaces used in Mississippian and Caddo imagery differ in noteworthy ways that are manifest in the associated material culture assemblages, and bifaces from Caddo sites more readily articulate with the serpent dance, where the motif portrays the emergence of the Hero Twins from the Beneath World; each grasping an elliptical biface.

@@

Among the shell cup illustrations assembled by Phillips and Brown [-@RN2447] that characterize the material culture at Spiro Mound is a depiction of elliptical bifaces conveyed as elements of the serpent/snake dance rather than elements of the eagle/bird dance. Snakes occupy an important place in Caddo lore where the Snake-Woman was the bringer of seeds to the Caddo, and no one, especially children, was permitted to point toward or touch the plants as they grow, since they belong to her until they are ripe [@RN11189, 18].

It is plausible that in the case of the serpent/snake dance, biface **shape** may convey a relevant---and ceremonially important---representation related to extant morphological differences expressed in local elapid and viper species. Differences in biface **size** may articulate with the size of a specific species, as some snakes are smaller than others (i.e. _Sistrurus miliarius streckeri_ or western pygmy rattlesnake). Alternatively, **size** differences might also depict allometric trends associated with the development and maturation of distinct species, as well as sexual size dimorphism.

@@

1.   Elliptical bifaces found at Millsap Cache were manufactured using Kay County Chert [@RN11461], while those from Jowell Farm were manufactured using Edwards Chert. Differences in stone tool shape as a function of raw material have been reported for Perdiz arrow points in the region [@RN9364]; however, others have demonstrated that the attribution of morphological variability to raw material differences is an assumption that warrants regular testing, suggesting that primacy should not be assumed [@RN8442]. Thus, a comparison of between-site morphology was included to assess elliptical biface morphology as a function of raw material.
  a)    The biogeography of elapid and viper species differs within the ancestral Caddo area, where elapid populations occur south of the Red River, excluding them from the geography surrounding Millsap Cache. If elliptical biface morphology was intended to represent local elapid and viper populations, then elliptical bifaces from these sites will differ due to differences in the constituent communities of elapids and vipers.

@@

In the case of ceremonial bifaces, the intentional use of parallel oblique flake scars may have been meant to convey motion, where flake scars on one face represent muscle constriction along one edge and muscle dilation on the other, and where that pattern is inverted on the opposite face. Thus, in intentionally flipping between the two faces---a potential component of the serpent/snake dance---dancers could convey motion using a static implement. The parallel oblique flake scars closely mimic the parallel scales that occur beneath both elapids and vipers, which shift in similar ways to generate motion.
